---
title: "Student Loan Defaults"
author: "Bo Suzow"
date: "July 30, 2018"
output: 
  html_document:
    keep_md: TRUE
---

# Introduction

Student loan debt (SLD) total in the US reached a staggering number in 1.5 trillion dollars, borrowed by 44 million people. It is `r round(1480/620,1)` times larger than the total of credit card debt.  The SLD total in 2008 was 640 billion dollars [which ballooned to 1.2 trillion by 2015](https://www.politifact.com/truth-o-meter/statements/2015/aug/14/jeb-bush/jeb-bush-student-loan-debt-has-doubled-under-obama/).  As the debt ever increases, what is the trend in its default rates?  Are they going up or held at steady rates?  

The U.S. Department of Education (USDOE) released [official cohort default rates for 3 cohorts](https://www2.ed.gov/offices/OSFAP/defaultmanagement/cdr.html#table) (2012, 2013 and 2014) in September 2017. This report analyzes the default data to identify any patterns and/or trend nationwide and by state. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(tidyr)
library(readxl)
```

# Data Load

## Student Loan Defaults (SLD)

The data set is publicly available [here](https://www2.ed.gov/offices/OSFAP/defaultmanagement/peps300.xlsx).  The details on the data dictionary (explanation of each column/field) is found [here](https://www2.ed.gov/offices/OSFAP/defaultmanagement/instructions.html).  While USDOE released 5 Default Reports, this analysis *focuses on the 2014 cohort* of Report 300 which reports "FY 2014, FY 2013, and FY 2012 official cohort default rates published for schools participating in the Title IV student financial assistance programs."

## College Scorecard (CS)

<<<<<<<<< To be revised with the ref of 10-14 data >>>>>>>>>>>>>>>>>

This data set is a part of a [data bundle](https://ed-public-download.app.cloud.gov/downloads/CollegeScorecard_Raw_Data.zip) published by the US Department of Education.  This analysis utilizes the data pertinent to the 2014-15 academic year cohort (`MERGED2014_15_PP.csv`). The full data documentation is found [here](https://collegescorecard.ed.gov/assets/FullDataDocumentation.pdf).


```{r data_load}

# Load the student loan default data set.

defaults <- read_excel("peps300.xlsx")
defaults <- tbl_df(defaults)

# Load the college scorecard data set for academic year 14-15

scorecard14-15<- read.csv("MERGED2014_15_PP.csv")
scorecard14-15 <- tbl_df(scorecard14-15)

# Load the college scorecard data set for academic year 10-11

scorecard10-11<- read.csv("MERGED2010_11_PP.csv")
scorecard10-11 <- tbl_df(scorecard10-11)


```

***
# Data Wrangling

The default data set (`defaults`) and the college scorecard data set (`scorecard`) will be joined on OPEID. This study will be conducted on the institutions represented in `defaults`.

## Strategy

- Clean up Student Loan Default (SLD) data set (`defaults`).
    - Standardize column names.
    - Inspect presence of missing value.
    - Determin missing value treatment strategy if applicable.
    - Select variables relevant to this study.
    
- Join SLD and the College Scorecard (CS) data set (`scorecard`). 
    - Review the colleges presented in multiple rows (example: OPEID 23328) and roll them up, maintaining contextual meanings data provides.
    
- Clean up `scorecard`.
    - Repeat the same steps applied to SLD.
       - The string `NULL` is used for missing value.  

### Student Loan Default Data Set - `defaults`

The dataset features 4712 schools participating in the Title IV programs. 29 variables/columns are provided in the dataset.  Let's rename them for easiness of the processing ensued.  

```{r}

names(defaults)=c("id","name","addr","city","state","state_name","zip","zipext","prog_len","school_type",
                  "yr1","num1","denum1","drate1","prate1",
                  "eth_code","program","congress_dist","region",
                  "yr2","num2","denum2","drate2","prate2",
                  "yr3","num3","denum3","drate3","prate3")
                  
```

Are there any columns with missiong values?

```{r}

names(defaults)[colSums(is.na(defaults))>0]

```

No missing value strategy is needed as these columns will not be used in the analysis. 

***

There are a few categorical (aka factor) variables in the dataset -- State (state), Program Length (prog_len), School Type (school_type), Program, Ethnic Code (eth_code).
Let's convert them to factor variables (categorical variables are called factor variables in R) with proper labels (descriptions of the codes).

```{r}

#-----------
# State
#-----------

defaults <- defaults %>% mutate(state = factor(state))

#-------------------
# Program Length 
#-------------------

prog_len_list = c(0:12)
prog_len_descs = c("Short Term (300-599 hours)",
                   "Graduate/Professional (>=300 hours)", 
                   "Non-Degree (600-899 hours)",
                   "Non-Degree 1 Year (900 - 1799 hours)",
                   "Non-Degree 2 Year (1800-2699 hours)",
                   "Associate's Degree","Bachelor's Degree",
                   "First Professional Degree",
                   "Master's Degree or Doctor's Degree",
                   "Professional Certification",
                   "Undergraduate (Previous Degree Required",
                   "Non-Degree 3 Plus Years (>= 2700 hours)",
                   "Two-Year Transfer")
defaults <- defaults %>% mutate(prog_len = factor(prog_len,levels=prog_len_list,
                                                  labels=prog_len_descs))
#-------------
# School Type
#-------------

school_type_list = c(1:3,5:7)
school_type_descs = c("Public",
                   "Private, Nonprofit", 
                   "Proprietary",
                   "Foreign public",
                   "Foreign private",
                   "Foreign For-Profit")
defaults <- defaults %>% mutate(school_type = factor(school_type,levels=school_type_list,
                                                  labels=school_type_descs))
#--------
# Region
#--------

region_list = c(1:11)
region_descs = c("01","02","03","04","05","06","07","08","09","10","11")

defaults <- defaults %>% mutate(region = factor(region,levels=region_list,
                                             labels=region_descs))

```


### Colleague Scorecard Data Set - `scorecard`

As there are a large number (1700+) of variables in `scorecard` data set, the [data element list](https://collegescorecard.ed.gov/assets/CollegeScorecardDataDictionary.xlsx) has been reviewed. Using the [data documentation](https://collegescorecard.ed.gov/assets/FullDataDocumentation.pdf) as guidelines, the variables are selected in the following code block:

```{r}

sc.net <- scorecard %>% select(OPEID6,INSTNM,STABBR,NUMBRANCH,CONTROL,PREDDEG,
                               REGION,LOCALE,LATITUDE,LONGITUDE,CCBASIC,
                               CCUGPROF,CCSIZSET,RELAFFIL,ADM_RATE_ALL,
                               DISTANCEONLY,UGDS,UG,
                               UGDS_WHITE,UGDS_BLACK,UGDS_HISP,UGDS_ASIAN,
                               CURROPER,
                               NPT4_PUB,NPT4_PRIV,
                               NUM4_PUB,NUM4_PRIV,
                               TUITFTE,INEXPFTE,AVGFACSAL,
                               PFTFAC,
                               PCTPELL,
                               C150_4,
                               RET_FT4,
                               PCTFLOAN,
                               UG25ABV,
                               CDR3,
                               PAR_ED_PCT_1STGEN,
                               DEP_INC_AVG,
                               IND_INC_AVG,
                               DEBT_MDN,
                               GRAD_DEBT_MDN,
                               WDRAW_DEBT_MDN,
                               FAMINC,
                               MD_FAMINC,
                               POVERTY_RATE,
                               MN_EARN_WNE_P10,
                               MD_EARN_WNE_P10,
                               CDR3_DENOM,
                               T4APPROVALDATE)

```



***
# Quick counts and plots
## School Counts by Categorical Variables/Columns

```{r}

sum_state <- defaults %>% group_by(state) %>% summarize(n())
sum_prog_len <- defaults %>% group_by(prog_len) %>% summarize(n())
sum_school_type <- defaults %>% group_by(school_type) %>% summarize(n())
sum_program <- defaults %>% group_by(program) %>% summarize(n())
sum_eth_code <- defaults %>% group_by(eth_code) %>% summarize(n())

```

***


```{r}

```

# Data Exploration

```{r}

```

# Data Plots

```{r}

```




