---
title: "Student Loan Defaults"
author: "Bo Suzow"
date: "July 30, 2018"
output: 
  html_document:
    keep_md: TRUE
---

# Introduction

Student loan debt (SLD) total in the US reached a staggering number in 1.5 trillion dollars, borrowed by 44 million people. It is `r round(1480/620,1)` times larger than the total of credit card debt.  The SLD total in 2008 was 640 billion dollars [which ballooned to 1.2 trillion by 2015](https://www.politifact.com/truth-o-meter/statements/2015/aug/14/jeb-bush/jeb-bush-student-loan-debt-has-doubled-under-obama/).  

The US Department of Education publishes the College Scorecard data to help the public to make informed decisions about investments in higher education.  The data features large amount of metrics including default rates and is organized by academic year. In this report, we will focus on the 2014-15 scorecard data.  The goals of this report are:

- Explore the data to ascertain some trends the data is telling us.
- Find out strong predictors affecting default rates.
- Build a prediction model for default rates.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(tidyr)
library(readxl)
```

# Data Load

This data set is a part of a [data bundle](https://ed-public-download.app.cloud.gov/downloads/CollegeScorecard_Raw_Data.zip) published by the US Department of Education.  This analysis utilizes the data pertinent to the 2014-15 academic year cohort (`MERGED2014_15_PP.csv`). The full data documentation is found [here](https://collegescorecard.ed.gov/assets/FullDataDocumentation.pdf).


```{r data_load, cache=TRUE}

# Load the college scorecard data set for academic year 14-15

scorecard1415<- read.csv("MERGED2014_15_PP.csv",na.strings=c("NULL","PrivacySuppressed"))
scorecard1415 <- tbl_df(scorecard1415)


```

***
# Data Wrangling

## Strategy

- Clean up the data set.
    - Select variables relevant to this study.
    - Standardize column names if necessary.
    - Inspect presence of missing value.
    - Determin missing value treatment strategy if applicable.
    - Transform data types appropriately.


## Select relevant variables 

As there are a large number (1700+) of variables in `scorecard` data set, the [data element list](https://collegescorecard.ed.gov/assets/CollegeScorecardDataDictionary.xlsx) has been reviewed. Using the [data documentation](https://collegescorecard.ed.gov/assets/FullDataDocumentation.pdf) as guidelines, the variables are selected in the following code block:


```{r selectVar1}

sc1415.all <- scorecard1415 %>% select(OPEID6,INSTNM,STABBR,NUMBRANCH,CONTROL,PREDDEG,
                               REGION,
                               LOCALE,LATITUDE,LONGITUDE,CCBASIC,
                               CCUGPROF,CCSIZSET,RELAFFIL,
                               ADM_RATE_ALL,
                               DISTANCEONLY,UGDS,UG,
                               UGDS_WHITE,UGDS_BLACK,UGDS_HISP,UGDS_ASIAN,
                               CURROPER,
                               NPT4_PUB,NPT4_PRIV,
                               NUM4_PUB,NUM4_PRIV,
                               TUITFTE,INEXPFTE,AVGFACSAL,
                               PFTFAC,
                               PCTPELL,
                               C150_4,
                               RET_FT4,
                               PCTFLOAN,
                               UG25ABV,
                               CDR2,
                               CDR3,
                               PAR_ED_PCT_1STGEN,
                               DEP_INC_AVG,
                               IND_INC_AVG,
                               DEBT_MDN,
                               GRAD_DEBT_MDN,
                               WDRAW_DEBT_MDN,
                               FAMINC,
                               MD_FAMINC,
                               POVERTY_RATE,
                               MN_EARN_WNE_P10,
                               MD_EARN_WNE_P10,
                               CDR3_DENOM
)
                               

```


The `CDR3` column reports the default rate. Hence, the observations with missing values in the column cannot be used for this analysis.  They are removed. 

```{r isNA}
 
sc1415 <- sc1415.all %>% filter(!is.na(CDR3))

```

It's observed that missing values are rather concentrated on the schools classified as stand alone graduate institutions. Let's remove them from the analysis.

```{r removeGradInst}
sc1415 <- sc1415 %>% filter(PREDDEG!=4)
```


## Missing Values

Even after removing the grad school rows, are there any columns with missiong values?

```{r colWithMVs}

names(sc1415)[colSums(is.na(sc1415))>0]


```

44 of `r ncol(sc1415)` columns still have missing values.  Let's compute their proportions and select those under 10%.

```{r naProportion}
naProportion <- apply(sc1415,2,function(x){sum(is.na(x))/nrow(sc1415)})
naProportion[naProportion<.1]


```


Let's confirm that all `CDR3` and `CDR3_DENOM` values are numeric.

```{r checkCDR3}

sum(!is.numeric(sc1415$CDR3))
sum(!is.numeric(sc1415$CDR3_DENOM))

```

## Variable selection

Now select the variables with less than 10% missing value. Further clean up by eliminating the rows with missing values (Caveat: This is a simpliest approach of treating missing values.  This may have to be revisited in pursuit of a more efficient strategy.)

(Qu: a better way to select the variables of interest without hard coding, but utilizing the output from the `naProportion` code chunk?)

```{r removeAllMVs}
sc1415 <- sc1415 %>% select(OPEID6, INSTNM, STABBR, NUMBRANCH, CONTROL, PREDDEG, REGION,
                             DISTANCEONLY, TUITFTE, INEXPFTE, 
                             PCTPELL, PCTFLOAN, CDR3, PAR_ED_PCT_1STGEN,
                             DEP_INC_AVG, IND_INC_AVG, 
                             DEBT_MDN, GRAD_DEBT_MDN, WDRAW_DEBT_MDN,
                             FAMINC, MD_FAMINC, CDR3_DENOM)

sc1415.net <- sc1415[complete.cases(sc1415),]


names(sc1415)[colSums(is.na(sc1415.net))>0]

```

The total number of rows has reduced from `r nrow(sc1415)` to `r nrow(sc1415.net)`. 
The summary statistics of default rates (`CDR3`) are comparable amongst the three data sets -- with graduate schools present, without grad schools, and without missing values.  We will move ahead with the the `sc1415.net` data frame for the analysis.  Please note that the unclassified institution (`PREDDEG = 0`) have been removed at the missing-value row elimination.

(Qu: Should box plots be added to confirm this point?)

```{r summaryCDR3}

summary(sc1415.all$CDR3)

summary(sc1415$CDR3)

summary(sc1415.net$CDR3)

```


## Converting to factor variables

Convert `CONTROL`, `PREDDEG`, `DISTANCEONLY`, and `REGION` to factor variables.

```{r facVars}

control_list = c(1:3)
control_descs = c("Public",
                  "Private nonprofit",
                  "Private for-profit")

sc1415.net <- sc1415.net %>% mutate(CONTROL = factor(CONTROL,levels=control_list,
                                             labels=control_descs))
# 1	Public
# 2	Private nonprofit
# 3	Private for-profit


preddeg_list = c(1:3)
preddeg_descs = c(
                  "Certificate",
                  "Associate's",
                  "Bachelor's"
                  )

sc1415.net <- sc1415.net %>% mutate(PREDDEG = factor(PREDDEG,levels=preddeg_list,
                                             labels=preddeg_descs))

# 0	Not classified	IPEDS  -- not included in the analysis
# 1	Predominantly certificate-degree granting	
# 2	Predominantly associate's-degree granting	
# 3	Predominantly bachelor's-degree granting	
# 4	Entirely graduate-degree granting -- not included in the analysis	


distanceonly_list = c(0:1)
distanceonly_descs = c("Not Online-Ed Only",
                  "Online-Ed Only")

sc1415.net <- sc1415.net %>% mutate(DISTANCEONLY = factor(DISTANCEONLY,levels=distanceonly_list,
                                             labels=distanceonly_descs))

#0	Not distance-education only
#1	Distance-education only


region_list = c(0:9)
region_descs = c("U.S. Service Schools",
                 "New England",
                 "Mid East",
                 "Great Lakes",
                 "Plains",
                 "Southeast",
                 "Southwest",
                 "Rocky Mtn",
                 "Far West",
                 "Outlying Areas")

sc1415.net <- sc1415.net %>% mutate(REGION = factor(REGION,levels=region_list,
                                             labels=region_descs))


# 0	U.S. Service Schools	
# 1	New England (CT, ME, MA, NH, RI, VT)		
# 2	Mid East (DE, DC, MD, NJ, NY, PA)		
# 3	Great Lakes (IL, IN, MI, OH, WI)		
# 4	Plains (IA, KS, MN, MO, NE, ND, SD)		
# 5	Southeast (AL, AR, FL, GA, KY, LA, MS, NC, SC, TN, VA, WV)		
# 6	Southwest (AZ, NM, OK, TX)		
# 7	Rocky Mountains (CO, ID, MT, UT, WY)		
# 8	Far West (AK, CA, HI, NV, OR, WA)		
# 9	Outlying Areas (AS, FM, GU, MH, MP, PR, PW, VI)		


```

# Exploratory Data Analysis (WIP)


## Quick Descriptive Statistics

The total number of students who entered repayment between October 2013 and September 2014 is `r round(sum(sc1415$CDR3_DENOM)/1000000,1)` million. Of these, `r sum(sc1415.net$CDR3_DENOM)` are the students from the certificate, associate's or bachelor's degree programs.  Their average default rate is `r mean(sc1415.net$CDR3)` with the standard deviation of `r sd(sc1415.net$CDR3)`.

Total count, mean and standard deviation of default rates (`CDR3`) by `CONTROL` - ownership type.

```{r}
table(sc1415.net$CONTROL)
sc1415.net %>% group_by(CONTROL) %>% summarize(mean(CDR3),sd(CDR3))
```

By `PREDDEG`- predominant degree awarded.

```{r}
table(sc1415.net$PREDDEG)
sc1415.net %>% group_by(PREDDEG) %>% summarize(mean(CDR3),sd(CDR3))

```

By `DISTANCEONLY` - whether distance education only or not.

```{r}
table(sc1415.net$DISTANCEONLY)
sc1415.net %>% group_by(DISTANCEONLY) %>% summarize(mean(CDR3),sd(CDR3))
```

By `REGION` - geographical location.

```{r}
table(sc1415.net$REGION)
sc1415.net %>% group_by(REGION) %>% summarize(mean(CDR3),sd(CDR3))
```

Summaries of numerical variables.

```{r}

sc1415.net %>% group_by(CONTROL) %>% summarize(mean(CDR3),sd(CDR3))

```



Let's draw some histograms.


```{r descStatPlots}

ggplot(sc1415.net,aes(x=CDR3)) + 
   geom_histogram(binwidth=.05,alpha=.5) +
   xlab("Default Rate")

ggplot(sc1415.net,aes(x=CDR3, fill=CONTROL)) + 
   geom_histogram(binwidth=.05) +
   facet_grid(CONTROL~.) +
   xlab("Default Rate")

ggplot(sc1415.net,aes(x=CDR3, fill=PREDDEG)) + 
   geom_histogram(binwidth=.05) +
   facet_grid(PREDDEG~.) +
   xlab("Default Rate")

ggplot(sc1415.net,aes(x=CDR3, fill=DISTANCEONLY)) + 
   geom_histogram(binwidth=.05) +
   facet_grid(DISTANCEONLY~.) +
   xlab("Default Rate")

ggplot(sc1415.net,aes(x=CDR3, fill=REGION)) + 
   geom_histogram(binwidth=.05) +
   facet_grid(REGION~.) +
   xlab("Default Rate")



```


## Scatter Plots

Numeric variables vs default rate by school ownership type

```{r}

# The median debt for students who have completed vs default rate by school ownership

ggplot(sc1415.net,aes(x=GRAD_DEBT_MDN,y=CDR3, col=CONTROL)) +
   geom_point(alpha=.3) +
   facet_grid(CONTROL~.) +
   geom_smooth(col="black")


# The median debt for students who have withdrawn vs default rate by school ownership

ggplot(sc1415.net,aes(y=CDR3,x=WDRAW_DEBT_MDN, col=CONTROL)) +
   geom_point(alpha=.3) +
   facet_grid(CONTROL~.) +
   geom_smooth(col="black")


# Percentage first-generation students vs default rate by school ownership

ggplot(sc1415.net,aes(y=CDR3,x=PAR_ED_PCT_1STGEN, col=CONTROL)) +
   geom_point(alpha=.3) +
   facet_grid(CONTROL~.) +
   geom_smooth(col="black")


# Percent of all undergraduate students receiving a federal student loan

ggplot(sc1415.net,aes(y=CDR3,x=PCTFLOAN, col=CONTROL)) +
   geom_point(alpha=.3) +
   facet_grid(CONTROL~.) + 
   geom_smooth(col="black")

```

A numeric variable XYZ vs default rate by predominant degree type

```{r}

# The median debt for students who have completed vs default rate by degree type

ggplot(sc1415.net,aes(y=CDR3,x=GRAD_DEBT_MDN, col=PREDDEG)) +
   geom_point(alpha=.3) +
   facet_grid(PREDDEG~.) +
   geom_smooth(col="black")


# The median debt for students who have withdrawn vs default rate by degree type

ggplot(sc1415.net,aes(y=CDR3,x=WDRAW_DEBT_MDN, col=PREDDEG)) +
   geom_point(alpha=.3) +
   facet_grid(PREDDEG~.) +
   geom_smooth(col="black")


# Percentage first-generation students vs default rate by degree type

# try geom_jitter()


ggplot(sc1415.net,aes(y=CDR3,x=PAR_ED_PCT_1STGEN, col=PREDDEG)) +
   geom_point(alpha=.3) +
   facet_grid(PREDDEG~.) +
   geom_smooth(col="black")



```


A numeric variable XYZ vs default rate by region

```{r}

# The median debt for students who have completed vs default rate by region

ggplot(sc1415.net,aes(y=CDR3,x=GRAD_DEBT_MDN, col=REGION)) +
   geom_point(alpha=.3) +
   facet_grid(REGION~.) +
   geom_smooth(col="black")

# The median debt for students who have withdrawn vs default rate by region

ggplot(sc1415.net,aes(y=CDR3,x=WDRAW_DEBT_MDN, col=REGION)) +
   geom_point(alpha=.3) +
   facet_grid(REGION~.) +
   geom_smooth(col="black")

# Percentage first-generation students vs default rate by region

# try geom_jitter()


ggplot(sc1415.net,aes(y=CDR3,x=PAR_ED_PCT_1STGEN, col=REGION)) +
   geom_point(alpha=.3) +
   facet_grid(REGION~.) +
   geom_smooth(col="black")


```